---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

Set some global options for processing the R markdown file
```{r setopts, echo=TRUE}
library(knitr, quietly=TRUE)
opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results="hide")

```

## Loading and preprocessing the data
Download zip-file from course repository and import included dataset into activity object. Results
are cached to avoid extensive downloading of the datafile.

```{r import data, cache=TRUE}
suppressPackageStartupMessages(library(tidyverse))
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
activity <- read_csv(unz(temp, "activity.csv"),
		     quote="\"",
		     col_types = cols(date=col_date(format="%Y-%m-%d")),
		     na = "NA")
unlink(temp)
```

## What is mean total number of steps taken per day?
To calculate the mean and median of the steps taken per day and to produce a histogram of these
values first the steps are summed up by day ignoring the NA values. From the resulting daily
summaries the mean and median is calculated.

```{r mean steps}
suppressPackageStartupMessages(library(tidyverse))
activity_by_day <- activity %>%
	group_by(date) %>%
	summarize(StepsSum = sum(steps, na.rm=TRUE))

meanSteps <- round(mean(activity_by_day$StepsSum),0)
medianSteps <- round(median(activity_by_day$StepsSum),0)
```

The (rounded) mean and median values for the steps taken per day are `r sprintf("%i", meanSteps)` and
`r sprintf("%i", medianSteps)`, respectively. A histogram including markes of the mean and median
is shown below.

```{r steps histogram, results="asis"}
p1 <- ggplot() + 
	geom_histogram(data = activity_by_day, aes(x=StepsSum), binwidth=1200) +
	geom_vline(aes(xintercept=meanSteps, colour="Mean")) + 
	geom_vline(aes(xintercept=medianSteps, colour="Median")) + 
	scale_x_continuous("Number of steps per day") +
	scale_colour_discrete("Summary Values") + 
	theme_bw()

print(p1)
```

## What is the average daily activity pattern?
To plot the average daily activity pattern the dataset is first grouped by interval calculating
the average numbers of steps. Subesquently the interval containing the maximum number of steps
is identified.

```{r summarize per interval}
suppressPackageStartupMessages(library(tidyverse))

activity_by_interval <- activity %>%
	group_by(interval) %>%
	summarize(MeanSteps = mean(steps, na.rm=TRUE))

max_activity_by_interval <- activity_by_interval %>%
	arrange(MeanSteps) %>%
	dplyr::top_n(1)

maxInterval <- max_activity_by_interval$interval

```

A plot of the average number of steps per interval is shown below highlighting interval
`r maxInterval` containing the maximum average step count.

```{r plot per interval, results="asis"}
p2 <- ggplot() + 
	geom_line(data = activity_by_interval, aes(x=interval, y=MeanSteps), group=1) +
	geom_vline(xintercept=maxInterval, colour="red") + 
	theme_bw()

print(p2)
```


## Imputing missing values

Some donwstream analyses might be affected by the missing ("NA") values for some day/interval
combinations in the dataset. In order to improve this the missing values are replaced by the mean
of the respective interval across the whole dataset.

```{r imputing missing values}
suppressPackageStartupMessages(library(tidyverse))

missingCount <- sum(is.na(activity$steps))
missingPercent <- round(missingCount/(missingCount+sum(!is.na(activity$steps)))*100,0)

activity_imputed <- activity %>%
	left_join(activity_by_interval) %>%
	mutate(steps_imputed = coalesce(steps,MeanSteps)) %>%
	select(steps_imputed, date, interval) %>%
	rename(steps = steps_imputed)

activity_by_day_imputed <- activity_imputed %>%
	group_by(date) %>%
	summarize(StepsSum = sum(steps, na.rm=TRUE))

meanStepsImputed <- round(mean(activity_by_day_imputed$StepsSum),0)
medianStepsImputed <- round(median(activity_by_day_imputed$StepsSum),0)

```

In total the dataset included `r missingCount` NA values for the steps taken which corresponds to 
`r missingPercent`% of the whole dataset. Using the imputation method described above the mean
changes to `r sprintf("%i", meanStepsImputed)` (from `r sprintf("%i", meanSteps)`) while the
median changes to `r sprintf("%i", medianStepsImputed)` (from `r sprintf("%i", medianSteps)`).

Below the histogram created from the imputed dataset is shown together with the highlighted 
mean and median values. Note that for the imputed dataset mean and median values are identical!

```{r steps histogram 2, results="asis"}
p3 <- ggplot() + 
	geom_histogram(data = activity_by_day_imputed, aes(x=StepsSum), binwidth=1200) +
	geom_vline(aes(xintercept=meanStepsImputed, colour="Mean")) + 
	geom_vline(aes(xintercept=medianStepsImputed, colour="Median")) + 
	scale_x_continuous("Number of steps per day") +
	scale_colour_discrete("Summary Values") + 
	theme_bw()

print(p3)
```


## Are there differences in activity patterns between weekdays and weekends?

For the analysis of different patterns first an additional variable "day" is created containing the
values "Weekday" or "Weekend" calculated from the "date" column. Then a panel plot is created
showing the average number of steps per interval for weekdays and weekends respectively. This plot
shows that while the overall pattern looks similar the large spike around interval 800 is much more
present on weekdays than on the weekends. 

```{r weekday patterns}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(timeDate))

activity_by_interval_and_day  <- activity_imputed %>%
	mutate(day = case_when(isWeekend(date) ~ "Weekend",
			       !isWeekend(date) ~ "Weekday")) %>%
	group_by(interval, day) %>%
	summarize(MeanSteps = mean(steps, na.rm=TRUE))

```

```{r plot per interval and weekday type, results="asis"}
p4 <- ggplot() + 
	geom_line(data = activity_by_interval_and_day, aes(x=interval, y=MeanSteps, group=day)) +
	facet_wrap(~day, ncol=1) +
	theme_bw()

print(p4)
```

